Stochastic Approximation Monte Carlo with a Dynamic Update Factor

### 2019-11-15 papers/histogram/sad.tex
edits on LJ section

\begin{figure}
  \includegraphics[width=\columnwidth]{figs/lj-cv}
  \caption{The heat capacity of the LJ31 cluster after $10^{12}$ moves
    as calculated with each method.}
  \label{fig:lj-cv}
\end{figure}

@@

In order to explore both the bin-size behavior and low-temperature
effectiveness of our algorithm, we chose to study the 31-atom cluster
(LJ31) because it is the smallest Lennard-Jones cluster that exhibits broken ergoidcity
and a solid-solid phase transition at a low temperature~\cite{doye1998thermodynamics, martiniani2014superposition}.
Clusters of Lennard-Jones atoms have been frequently used for testing
Monte Carlo algorithms~\cite{wales1997global, neirotti2000phase, frantsuzov2005size, mandelshtam2006multiple}.  The LJ31
cluster is commonly used for testing convergence of algorithms
because it features a two-funnel energy landscape with a significant
barrier between the two low-energy states~\cite{calvo2000entropic, calvo2000phase, poulain2006performances, martiniani2014superposition}.
In particular, the LJ31 cluster features a solid-solid phase
transition at a temperature of around $T \approx 0.027\epsilon/k_B$ that is
challenging to converge~\cite{mandelshtam2006multiple}.  Poulain \emph{et al.}
tested a number of variants of WL on LJ31 and found
\davidsays{summarize}~\cite{poulain2006performances}.

\jpcom{This \P\ should be commented once summarized.
Poulain \emph{et al.} found very poor convergence for a variety of
modified WL methods on the LJ31
cluster~\cite{poulain2006performances}.  Sadly, they don't mention the
energy range or bin size that was used in their simulation, so their
results cannot be reproduced.  They do mention that they use 1000
energy bins, which if they used an energy range of
$[-133.58\epsilon,0]$ would correspond to a bin size of $\Delta
E\approx 0.13\epsilon$, which we find to be too large to accurately
predict the heat capacity peak at $0.03\epsilon/k_B$.  Looks like they
only ran $\sim 10^{10}$ moves. We can't converge in that few moves either.}

Figure~\ref{fig:lj-cv} shows the heat capacity of the LJ31 cluster as a function
of temperature.  There are two peaks, a large peak around $T \approx
0.3\epsilon/k_B$? corresponding to melting temperature, and a small peak around
$T \approx 0.027\epsilon/k_B$ for the solid-solid transition from a
Mackay-to-anti-Mackay (M $\rightarrow$ aM) transition~\cite{doye1997thermally,
hendy2001molecular, mandelshtam2006multiple}.  We keep our atoms in a spherical
box with radius 2.5$\sigma$, as is common in the
literature~\cite{poulain2006performances, mandelshtam2006multiple}.  We find
that using a larger box (with radius $5\sigma$) has a large effect on the
melting peak, and a smaller but still significant effect on the solid-solid
transition.  Since our focus is on the convergence rate, we focused on the
smaller box size.  We further restricted our simulations to negative energies,
as the microcanonical energy at infinite temperature is rather high for such a
small box.

\begin{figure}
  \includegraphics[width=\columnwidth]{figs/lj-cv-error}
  \caption{The maximum error in the heat capacity of the LJ31 cluster
    in the temperature range from 0.01~$\epsilon/k_B$ to
    0.1~$\epsilon/k_B$.  Note that the value of $C_V$ ranges from
    90$k_B$ to 120$k_B$, so the smallest (maximum) errors seen are
    around 1\%.  All WL methods shown were run with a histogram bin
    width of $\Delta E=0.01\epsilon$.  Their convergence at other bin
    widths was similar to that of SAD. }
  \label{fig:lj-cv-error}
\end{figure}

\jpcom{this is a basically a repeat of above...

The heat capacity changes significantly even at these low temperatures
when the size of the cavity is increased.  We use a 2.5 cavity because
``everyone else'' does~\cite{poulain2006performances}.  For purposes
of testing the convergence of the algorithm, we don't need to solve
for the true heat capacity.  Furthermore, the heat capacity at fixed
volume is not the most meaningful quantity to predict for a cluster.
}

Dependence on bin width.  Figure~\ref{fig:lj-cv-error} shows that the
heat capacity converges just as well for bin widths varying over two
orders of magnitude, although for bins as large as $\Delta
E=0.1\epsilon$ the error in the heat capacity is significant, as is
apparent in Fig~\ref{fig:lj-cv}.  The entropy itself converges much
more slowly for a smaller $\Delta E$, but when summing over energies
the statistical fluctuations in the entropy cancel. WL and $1/t$-WL
show similar behavior, so this is not new, but is encouraging.

Specifying temperature range rather than energy range is beneficial.
Both WL and $1/t$-WL are incredibly sensitive to the minimum energy
chosen, as seen in Fig.~\ref{fig:lj-cv-error}.  Cite previous work
that showed them not converging because it chose too low a minimum
energy, proving that it is \emph{not} a trivial procedure to determine
an appropriate energy range.  Specifying the temperature range
\emph{does} matter! We had to do a whole bunch of $1/t$-WL
computations in order to determine what would be a sufficient
low-energy cutoff.

SAMC is a major pain.  The $t_0$ value needs to be adjusted when
$\Delta E$ is changed or when the energy range is changed.  The energy
range has the same downsides as the WL methods, although it manifests
differently, with failure to converge as the problem, rather than
failure for $\gamma_t$ to decrease.

>>>>

\begin{figure}
  \includegraphics[width=\columnwidth]{figs/lj-cv}
  \caption{The heat capacity of the LJ31 cluster after $10^{12}$ moves
    as calculated with each method.  Each simulation used a bin size
    of $\Delta E=0.01\epsilon$.  The dotted black line labeled
    ``converged result'' was a $1/t$-WL simulation with the energy
    constrained to a range $-133.53\epsilon\le E\le -110\epsilon$,
    all other simulations had a maximum allowed energy of zero.  The
    ``converged result'' therefore is only converged at low temperatures.}
  \label{fig:lj-cv}
\end{figure}

@@

In order to explore both the bin-size behavior and low-temperature
convergence of our algorithm, we chose to study the 31-atom cluster
(LJ31), which is the smallest Lennard-Jones cluster that exhibits
broken ergoidcity and a solid-solid phase transition at a low
temperature~\cite{doye1998thermodynamics,
  martiniani2014superposition}.  Clusters of Lennard-Jones atoms have
been frequently used for testing Monte Carlo
algorithms~\cite{wales1997global, neirotti2000phase,
  frantsuzov2005size, mandelshtam2006multiple}.  The LJ31 cluster
specifically is commonly used for testing convergence of algorithms
because it features a two-funnel energy landscape with a significant
barrier between the two low-energy states~\cite{calvo2000entropic,
  calvo2000phase, poulain2006performances,
  martiniani2014superposition}.  In particular, the LJ31 cluster
features a solid-solid phase transition at a temperature of around $T
\approx 0.027\epsilon/k_B$ that is challenging to
converge~\cite{mandelshtam2006multiple}.  Poulain \emph{et al.}
tested a number of variants of the WL algorithm on LJ31 and failed to
converge the spike in heat capacity corresponding to the solid-solid
phase transition~\cite{poulain2006performances}.

%% \jpcom{This \P\ should be commented once summarized.
%% Poulain \emph{et al.} found very poor convergence for a variety of
%% modified WL methods on the LJ31
%% cluster~\cite{poulain2006performances}.  Sadly, they don't mention the
%% energy range or bin size that was used in their simulation, so their
%% results cannot be reproduced.  They do mention that they use 1000
%% energy bins, which if they used an energy range of
%% $[-133.58\epsilon,0]$ would correspond to a bin size of $\Delta
%% E\approx 0.13\epsilon$, which we find to be too large to accurately
%% predict the heat capacity peak at $0.03\epsilon/k_B$.  Looks like they
%% only ran $\sim 10^{10}$ moves. We can't converge in that few moves either.}

Figure~\ref{fig:lj-cv} shows the heat capacity of the LJ31 cluster
over the temperature range from $k_BT=0.01\epsilon$ to
$k_BT=0.40\epsilon$.  There are two peaks, a large peak around $k_BT
\approx 0.3\epsilon$? corresponding to melting temperature, and a
small peak around $T \approx 0.027\epsilon/k_B$ for the solid-solid
transition from a Mackay-to-anti-Mackay (M $\rightarrow$ aM)
transition~\cite{doye1997thermally, hendy2001molecular,
  mandelshtam2006multiple}.  We constrain our atoms within a spherical
box of radius 2.5$\sigma$, as is common in the
literature~\cite{poulain2006performances, mandelshtam2006multiple}.
We note that using a larger box (with radius $5\sigma$) has a large
effect on the melting peak, and a smaller but still significant effect
on the solid-solid transition.  Since our focus is on the convergence
rate, we focused on the smaller box size.  We further restricted all
our simulations to negative energies.

\begin{figure}
  \includegraphics[width=\columnwidth]{figs/lj-cv-error}
  \caption{The maximum error in the heat capacity of the LJ31 cluster
    in the temperature range 0.01~$\epsilon\le k_BT \le 0.05\epsilon$.
    Note that the value of $C_V$ ranges from 90$k_B$ to 120$k_B$, so
    the smallest (maximum) errors seen are around 1\%.  All WL methods
    shown were run with a histogram bin width of $\Delta
    E=0.01\epsilon$.  Their convergence at other bin widths was
    similar to that of SAD. }
  \label{fig:lj-cv-error}
\end{figure}

In order to test the low-temperature convergence, we focused on the
heat capacity in the temperature range $0.01\epsilon \le k_BT \le
0.05\epsilon$, which contains the solid-solid phase transition.
%
We compute the heat capacity from the entropy using a standard
canonical analysis, which gives
\begin{align}
  C_V &= 31\times\frac32k_B + \frac1{k_BT^2}\left< (E- \langle
  E\rangle)^2\right>
  \label{eq:cv}
\end{align}
where the first term represents the kinetic energy contribution, and
the second term is the potential energy contribution, with canonical
averages using our microcanonical entropy to compute multiplicities.
%
For our reference heat capacity, we ran a $1/t$-WL simulation with the
energy constrained to a range $-133.53\epsilon\le E\le -110\epsilon$.
We chose this energy range through experimentation, as we found the
convergence rate to be very sensitive to the minimum energy chosen.

%% \jpcom{this is a basically a repeat of above...
%% The heat capacity changes significantly even at these low temperatures
%% when the size of the cavity is increased.  We use a 2.5 cavity because
%% ``everyone else'' does~\cite{poulain2006performances}.  For purposes
%% of testing the convergence of the algorithm, we don't need to solve
%% for the true heat capacity.  Furthermore, the heat capacity at fixed
%% volume is not the most meaningful quantity to predict for a cluster.
%% }
We ran simulations with bin widths of $\Delta E=0.1\epsilon$,
$0.01\epsilon$, and $0.001\epsilon$ using each algorithm we studied.
When examining the converged results, we saw that each bin size
measureably different heat capacity at low temperatures. The largest
bin size of $0.1\epsilon$ introducing a maximum error in heat capacity
of more than 20$k_B$ while the intermediate bin size of $0.01\epsilon$
introduced an error of less than $k_B$ when compared with the
converged result at our smallest bin size.  Therefore we chose to
compare our simulations primarily at the intermediate bin size.  We
used a bin size of $0.01\epsilon$ for our reference in all plots in
this paper (including the reference), except where otherwise
specified.

Figure~\ref{fig:lj-cv-error} shows the largest error in heat capacity
at low temperatures as a function of the number of Monte Carlo moves
performed.
%
We explored the convergence behavior of each algorithm as a function
of bin width, but show only the results for SAD.  Both the WL and
$1/t$-WL algorithms showed similar behavior to SAD.  In each of these
algorithms, the rate of convergence of the heat capacity is roughly
independent of the bin width.  In contrast (but not shown) the entropy
itself shows significantly larger statistical errors with smaller bin
sizes.  The canonical averaging process in Eq.~\ref{eq:cv} averages
out those statistical fluctuations, largely eliminating the penalty of
using a smaller bin size.

%% Dependence on bin width.  Figure~\ref{fig:lj-cv-error} shows that the
%% heat capacity converges just as well for bin widths varying over two
%% orders of magnitude, although for bins as large as $\Delta
%% E=0.1\epsilon$ the error in the heat capacity is significant, as is
%% apparent in Fig~\ref{fig:lj-cv}.  The entropy itself converges much
%% more slowly for a smaller $\Delta E$, but when summing over energies
%% the statistical fluctuations in the entropy cancel. WL and $1/t$-WL
%% show similar behavior, so this is not new, but is encouraging.

We found the convergence of the WL and $1/t$-WL to be incredibly
sensitve to the minimum energy chosen, complicating the process of
obtaining a workable result.  The known ground state energy of this
cluster is \davidsays{look up and cite}.  With a minimum energy of
$-133.58\epsilon$, our simulation did not converge in 10$^{12}$ moves.
With minimum energies of $-133.54\epsilon$, $-133.53\epsilon$, and
$-133.52\epsilon$, the simulations do seem to converge, although the
lowest energy case was significantly slower and never reached as small
an error.  This result highlights the advantage of being able to
specify a temperature range rather than an energy range when computing
properties as a function of temperature.  It is not only more
convenient, but also far more efficient due the the fact that only a
single simulation need be performed.

%% Specifying temperature range rather than energy range is beneficial.
%% Both WL and $1/t$-WL are incredibly sensitive to the minimum energy
%% chosen, as seen in Fig.~\ref{fig:lj-cv-error}.  Cite previous work
%% that showed them not converging because it chose too low a minimum
%% energy, proving that it is \emph{not} a trivial procedure to determine
%% an appropriate energy range.  Specifying the temperature range
%% \emph{does} matter! We had to do a whole bunch of $1/t$-WL
%% computations in order to determine what would be a sufficient
%% low-energy cutoff.

Finally, we want to mention our results for SAMC, which are not shown
in our plots in order to reduce clutter.  With the right value for
$t_0$ and with the right energy range chosen, SAMC converges as
rapidly as either $1/t$-WL or SAD.  However, with a wrong $t_0$
parameter, the convergence rate is very poor.  In addition, the
optimal $t_0$ parameter depends on the bin width as well as the energy
range chosen, further complicating matters.

%% SAMC is a major pain.  The $t_0$ value needs to be adjusted when
%% $\Delta E$ is changed or when the energy range is changed.  The energy
%% range has the same downsides as the WL methods, although it manifests
%% differently, with failure to converge as the problem, rather than
%% failure for $\gamma_t$ to decrease.
